---
title: "CMI-PB 2nd challenge: Process prediction datasets similar to training dataset"
#format: html
#editor: visual
output:
  BiocStyle::html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    code_folding: show
    number_sections: false
date: "2023-12-21"
---

## Download all data files

We manually downloaded data files manually from the following repositories: 1) Raw prediction datasets: [CMI-PB https site](https://www.cmi-pb.org/downloads/cmipb_challenge_datasets/current/2nd_challenge/raw_datasets/). 2) Processed training datasets for ID mapping: [CMI-PB https site](https://www.cmi-pb.org/downloads/cmipb_challenge_datasets/current/2nd_challenge/processed_datasets/).

Alternatively, `RCurl` package can be used to download data files.

```{r}
#base_dir = "/home/pramod/Documents/GitHub/second-challenge-prediction-dataset-preparation/"
base_dir = "../"


## `codebase.R` installs required packages and all house keepinf functions
source(paste0(base_dir, "scripts/codebase.R"))


dir_data = paste0(base_dir, "data/")
dir_raw_prediction <- paste0(base_dir, "data/raw_datasets/prediction/")
dir_processed_prediction <- paste0(base_dir, "data/processed_datasets/prediction/")
```

## Read data

#### Read the training data object to associate each feature with its corresponding ID

```{r}
master_processed_training_data_obj = read_rds(paste0(dir_data, "master_processed_training_data.RDS"))

features_training_abtiter = rownames(master_processed_training_data_obj$abtiter_wide$batchCorrected_data)
features_training_pbmc_gene_expression = rownames(master_processed_training_data_obj$pbmc_gene_expression$batchCorrected_data)
features_training_plasma_cytokine_concentrations = rownames(master_processed_training_data_obj$plasma_cytokine_concentrations$batchCorrected_data)
features_training_pbmc_cell_frequency = rownames(master_processed_training_data_obj$pbmc_cell_frequency$batchCorrected_data)

```

#### Read subject and specimen ID mapping inforamtion

```{r}
d2022_subject <- read_tsv(paste0(dir_raw_prediction, "2022BD_subject.tsv"))
d2022_specimen <- read_tsv(paste0(dir_raw_prediction, "2022BD_specimen.tsv"))

## create new object subject_specimen
d2022_subject_specimen <- d2022_specimen %>%
  left_join(d2022_subject) %>%
  mutate(timepoint = planned_day_relative_to_boost)

```

#### Read experimental data: four assays

```{r warning=FALSE}
d2022_pbmc_abtiter <- read_tsv(paste0(dir_raw_prediction, "2022BD_plasma_ab_titer.tsv"))
d2022_pbmc_cell <- read_tsv(paste0(dir_raw_prediction, "2022BD_pbmc_cell_frequency.tsv"))
d2022_pbmc_gene <- read_tsv(paste0(dir_raw_prediction, "2022BD_pbmc_gene_expression.tsv"))
d2022_plasma_cytokine <- read_tsv(paste0(dir_raw_prediction, "2022BD_plasma_cytokine_concentration.tsv"))


```

## Process datasets

### abtiter data:

Identify feature overlaps between the prediction dataset and the training abtiter dataset; reformat the prediction abtiter data to align with the training data structure.

```{r}
#head(master_processed_training_data_obj$abtiter_wide$batchCorrected_data)

d2022_pbmc_abtiter_v1 = d2022_pbmc_abtiter %>%
  mutate(isotype_antigen = paste0(isotype, "_", antigen)) %>%
  filter(isotype_antigen %in% features_training_abtiter) %>%
  dplyr::select(specimen_id, isotype_antigen, MFI_normalised) %>%
  pivot_wider(names_from = specimen_id, values_from = MFI_normalised) 

d2022_pbmc_abtiter_v1_mat = d2022_pbmc_abtiter_v1 %>%
  column_to_rownames("isotype_antigen")

## Check training Vs prediction dataset dimentions
all(rownames(d2022_pbmc_abtiter_v1_mat) %in% rownames(master_processed_training_data_obj$abtiter_wide$batchCorrected_data))
```

### Gene expression data:

Identify feature overlaps between the prediction dataset and the training gene expression dataset; reformat the prediction gene expression data to align with the training data structure.

```{r}
#head(master_processed_training_data_obj$pbmc_gene_expression$batchCorrected_data)

d2022_pbmc_gene_v1 = d2022_pbmc_gene %>%
  filter(versioned_ensembl_gene_id %in% features_training_pbmc_gene_expression) %>% ## Filter
  dplyr::select(specimen_id, versioned_ensembl_gene_id, tpm) %>%
  pivot_wider(names_from = specimen_id, values_from = tpm) 

d2022_pbmc_gene_v1_mat = d2022_pbmc_gene_v1 %>%
  column_to_rownames("versioned_ensembl_gene_id")

## Check training Vs prediction dataset dimentions
all(rownames(d2022_pbmc_gene_v1_mat) %in% rownames(master_processed_training_data_obj$pbmc_gene_expression$batchCorrected_data))
```

### Cell Frequency Analysis:

Identify the median at baseline (day 0) in a manner similar to the training dataset. Then, identify feature overlaps between the prediction dataset and the training cell frequency dataset; subsequently, reformat the prediction cell frequency data to align with the training data structure.

```{r}
## Perform median normalization
cell_median_D0 <- d2022_pbmc_cell %>%
    left_join(d2022_subject_specimen) %>%
    filter(specimen_id %in% unique(d2022_subject_specimen[d2022_subject_specimen$planned_day_relative_to_boost == 0,]$specimen_id)) %>%
    group_by(dataset, cell_type_name)  %>%
    summarise(median = median(percent_live_cell, na.rm = T))

## Check if all cell_types have benn captured 
all(cell_median_D0$cell_type_name %in% d2022_pbmc_cell$cell_type_name)

d2022_pbmc_cell_normalized_pre <-  d2022_pbmc_cell  %>%
    left_join(d2022_subject_specimen) %>%
    left_join(cell_median_D0) %>%
    mutate(percent_live_cell_normalized = if_else(is.na(percent_live_cell) == T, NA, percent_live_cell/median))

## Reshape dataframe in wide format
d2022_pbmc_cell_normalized_pre_v1 <- d2022_pbmc_cell_normalized_pre  %>%
    filter(cell_type_name %in% features_training_pbmc_cell_frequency) %>% ## Filter
    dplyr::select(cell_type_name, specimen_id, percent_live_cell_normalized) %>%
    pivot_wider(names_from = "cell_type_name", values_from = percent_live_cell_normalized) 

d2022_pbmc_cell_normalized_pre_v1_mat = d2022_pbmc_cell_normalized_pre_v1 %>%
    column_to_rownames("specimen_id")%>%
    t() 

## Check training Vs prediction dataset dimentions
all(rownames(d2022_pbmc_cell_normalized_pre_v1_mat) %in% rownames(master_processed_training_data_obj$pbmc_cell_frequency$batchCorrected_data))


## Note: This step is optional and should be selected according to the specific needs of the modeling process. For the preparation of the training dataset, KNN imputation was utilized.

#d2022_pbmc_cell_normalized_pre_v1_mat_imputed = d2022_pbmc_cell_normalized_pre_v1_mat[rowMeans(is.na(d2022_pbmc_cell_normalized_pre_v1_mat)) < 1, ] %>%
#    as.matrix() %>%
#    impute.knn() %>%
#    .$data
```

### Plasma Cytokine Analysis:

Identify the median at baseline (day 0) in a manner similar to the training dataset. Then, identify feature overlaps between the prediction dataset and the training plasma cytokine dataset; subsequently, reformat the prediction plasma cytokine data to align with the training data structure.

```{r}
## Perform median normalization
cytokine_median_D0 <- d2022_plasma_cytokine %>%
    left_join(d2022_subject_specimen) %>%
    filter(specimen_id %in% unique(d2022_subject_specimen[d2022_subject_specimen$planned_day_relative_to_boost == 0,]$specimen_id)) %>%
    group_by(dataset, protein_id)  %>%
    summarise(median = median(protein_expression, na.rm = T))

d2022_plasma_cytokine_normalized_pre <-  d2022_plasma_cytokine  %>%
    left_join(d2022_subject_specimen) %>%
    left_join(cytokine_median_D0) %>%
    mutate(protein_expression_normalized = if_else(is.na(protein_expression) == T, NA, protein_expression/median))
  
## Reshape dataframe in wide format
d2022_plasma_cytokine_normalized_pre_v1 <- d2022_plasma_cytokine_normalized_pre  %>%
    filter(protein_id %in% features_training_plasma_cytokine_concentrations) %>% ## Filter
    dplyr::select(protein_id, specimen_id, protein_expression_normalized) %>%
    pivot_wider(names_from = "protein_id", values_from = protein_expression_normalized)

d2022_plasma_cytokine_normalized_pre_v1_mat <- d2022_plasma_cytokine_normalized_pre_v1  %>%
    column_to_rownames("specimen_id")%>%
    t() 

## Check training Vs prediction dataset dimentions
all(rownames(d2022_plasma_cytokine_normalized_pre_v1_mat) %in% rownames(master_processed_training_data_obj$plasma_cytokine_concentrations$batchCorrected_data))


## Note: This step is optional and should be selected according to the specific needs of the modeling process. For the preparation of the training dataset, KNN imputation was utilized.

#d2022_plasma_cytokine_normalized_pre_v1_mat_imputed = d2022_plasma_cytokine_normalized_pre_v1_mat[rowMeans(is.na(d2022_plasma_cytokine_normalized_pre_v1_mat)) < 1, ] %>%
#    as.matrix() %>%
#    impute.knn() %>%
#    .$data


  
```

### Save normalized data
#### Prepare the RDS object 

```{r}
master_processed_prediction_data <- list(
  
  subject_specimen = d2022_subject_specimen,
  abtiter = list(
              raw_from_database = d2022_pbmc_abtiter,
              processed_similar_to_training = d2022_pbmc_abtiter_v1_mat
          ),
  plasma_cytokine_concentrations = list(
              raw_from_database = d2022_plasma_cytokine,
              processed_similar_to_training = d2022_plasma_cytokine_normalized_pre_v1_mat
          ),
  pbmc_cell_frequency = list(
              raw_from_database = d2022_pbmc_cell,
              processed_similar_to_training = d2022_pbmc_cell_normalized_pre_v1_mat
          ),
  pbmc_gene_expression = list(
              raw_from_database = d2022_pbmc_gene,
              processed_similar_to_training = d2022_pbmc_gene_v1_mat
          )
  
)
```

#### Save the data in both RDS format and as individual TSV files for versatile use and accessibility
##### Note: Uncomment this section when you are ready to save the files.

```{r}
## Save RDS object

#saveRDS(master_processed_prediction_data, file = paste0(dir_processed_prediction, "master_processed_prediction_data.RDS"))

## Save individual data files

#write_tsv(d2022_subject_specimen, paste0(dir_processed_prediction, "subject_specimen.tsv"))
#write.table(d2022_pbmc_abtiter_v1_mat, 
#                file = paste0(dir_processed_prediction, "abtiter_processed_data.tsv"),  
#                sep = "\t", row.names = TRUE, quote = FALSE
#            )

#write.table(d2022_plasma_cytokine_normalized_pre_v1_mat, 
#                file = paste0(dir_processed_prediction, "plasma_cytokine_concentrations_processed_data.tsv"),  
#                sep = "\t", row.names = TRUE, quote = FALSE
#            )

#write.table(d2022_pbmc_cell_normalized_pre_v1_mat, 
#                file = paste0(dir_processed_prediction, "pbmc_cell_frequency_processed_data.tsv"),  
#                sep = "\t", row.names = TRUE, quote = FALSE
#            )

#write.table(d2022_pbmc_gene_v1_mat, 
#                file = paste0(dir_processed_prediction, "pbmc_gene_expression_processed_data.tsv"),  
#                sep = "\t", row.names = TRUE, quote = FALSE
#            )


```

## session_info()

```{r}
sessioninfo::session_info()
```
